Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-05 astronum_2019_MAESTROeX/astronum-maestroex.tex
changes to GPU section

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\maestroex}{{\sffamily MAESTROeX}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}
\newcommand{\starkiller}{{\sffamily Starkiller Microphysics}}
\newcommand{\amrexastro}{{\sffamily AMReX-Astro}}
\newcommand{\fboxlib}{{\sffamily FBoxLib}}

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. Its development is described in a series of papers: \cite{Almgren2006a,Almgren2006b,Almgren2008a,Zingale2009,Nonaka2010}. In \maestro, the system of low Mach number equations is solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE~\cite{vode}), and multigrid linear solvers for the pressure projection steps. Strang splitting~\cite{strang:1968} is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the pure Fortran 90 \fboxlib~software framework \cite{Zhang2016}; \maestroex~instead uses the C++/Fortran 90 \amrex~framework \cite{Zhang2019}.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. \castro~and \maestroex~share a common set of microphysics solvers provided by the \starkiller~library\footnote{\url{https://github.com/starkiller-astro/Microphysics}}. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. A prime example of this is the GPU port capability that has recently been implemented in \castro, and are currently in the process of being implemented in \maestroex. 
% \MarginPar{mention starkiller astrophysics}

@@

So far we have ported most of the source terms and hydrodynamics in \maestroex~to run on GPUs. GPU support for the linear solvers is becoming increasingly available from the \amrex~team.  We have found that one of the challenges moving forwards is dealing with the one-dimensional base state. Currently, each MPI process holds a copy of the entire base state. This is not a problem for small problems where the base state does not contain many cells compared to the number of cells in a single subgrid, however as the problem size and/or resolution gets larger, the size of the base state grows to the point that it is now significantly larger than the size of a single subgrid. Compared to CPUs, GPUs typically have less memory. For larger problems, it is therefore not possible to store the entire base state on the GPU. This significantly reduces the performance of the GPU parts of the code, as it will increase the amount of data will be copied back and forth between the CPU and GPU (a slow process). To alleviate this problem, we are looking at ways to reduce the amount of base state data that needs to be stored on the GPU at any one time. For example, in some routines we map the base state to the three-dimensional grid \textendash~for these, we can move this operation to the CPU so that the GPU does not require the base state. It may be necessary for us to develop a way to split up the base state, so that subgrids only receive the sections of the base state necessary for their calculations (rather than the base state for the entire problem domain).

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the three-dimensional \texttt{reacting\_bubble} problem on the OLCF Summit machine.
This problem uses a simple reaction network modelling $^{12}\mathrm{C} + {}^{12}\mathrm{C} \rightarrow {}^{24}\mathrm{Mg}$ that can be integrated on the GPUs using the CUDA Fortran port of VODE \citep{vode} described in \cite{Zingale2018}.
Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU.
Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.

>>>>

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\maestroex}{{\sffamily MAESTROeX}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}
\newcommand{\starkiller}{{\sffamily Starkiller Microphysics}}
\newcommand{\amrexastro}{{\sffamily AMReX-Astro}}
\newcommand{\fboxlib}{{\sffamily FBoxLib}}
\newcommand{\boxlib}{{\sffamily BoxLib}}

@@

Prior to \maestroex, we developed the low Mach number code \maestro. Like \maestroex, \maestro~is a block-structured adaptive mesh refinement (AMR) code for modelling low Mach number astrophysical codes. Its development is described in a series of papers: \cite{Almgren2006a,Almgren2006b,Almgren2008a,Zingale2009,Nonaka2010}. In \maestro, the system of low Mach number equations is solved using an explicit Godunov approach for the advection, a stiff ODE solver for the reactions (VODE~\cite{vode}), and multigrid linear solvers for the pressure projection steps. Strang splitting~\cite{strang:1968} is used to integrate the thermodynamic variables, a second order projection method to integrate the velocity subject to the divergence constraint, and a velocity splitting scheme to hydrodynamically evolve the base state. The original \maestro~code was developed using the Fortran 90 interface of the \boxlib~software framework \cite{Zhang2016}; \maestroex~instead uses the C++/Fortran 90 \amrex~framework \cite{Zhang2019}.

@@

\maestroex~is part of the \amrexastro~suite of open source adaptive mesh refinement hydrodynamics codes for astrophysical flows. Other codes in this family include \castro, an astrophysical radiation hydrodynamics simulation code \cite{Almgren2010}, and \nyx, an N-body hydrodynamics cosmological simulation code \cite{Almgren2013}. All three codes are developed using the \amrex~software framework, and share much in common in terms of their development, structure and numerical methods. \castro~and \maestroex~share a common set of microphysics solvers provided by the \starkiller~library\footnote{\url{https://github.com/starkiller-astro/Microphysics}}. Developers of the different codes work closely together (in fact many of the developers work on more than one of the codes), with their shared structure and framework meaning that new features developed in one code can easily be replicated in the others. A prime example of this is the GPU porting capability that has recently been implemented in \castro, and is currently in the process of being implemented in \maestroex. 
% \MarginPar{mention starkiller astrophysics}

@@

So far we have ported most of the source terms and hydrodynamics in \maestroex~to run on GPUs. GPU support for the linear solvers is becoming increasingly available from the \amrex~team.  We have found that one of the challenges moving forwards is dealing with the one-dimensional base state. Currently, each MPI process holds a copy of the entire base state. As the problem size and/or resolution gets larger, the size of the base state grows. Data transfers between the CPU and GPU are typically a slow process, so as the base state grows in size, the cost of these transfers becomes more significant and reduces the performance overall. As we are still in the process of porting our code to GPUs, there still remain some functions operating on the base state which run on the CPU. Every time one of these is called, if the necessary base state data is currently on the GPU, it is copied back to the CPU (and then back again next time it is required for a GPU-based calculation).

To alleviate this problem, we are looking at ways to reduce both the number and size of these data transfers. For example, in some routines we map the base state to the three-dimensional grid \textendash~for these, we can move this operation to the CPU so that the GPU does not require the entire 1d base state to be copied over. It may be necessary for us to develop a way to split up the base state, so that subgrids only receive the sections of the base state necessary for their calculations (rather than the base state for the entire problem domain).

As mentioned, our work porting \maestroex~to run on GPUs is still ongoing, however in Figure~\ref{fig:gpu_speedup} we show the speedup we have achieved so far for a number of individual functions that have been offloaded to GPU. The plot shows the average execution time of each function, recorded for the three-dimensional \texttt{reacting\_bubble} problem on the OLCF Summit machine.
This problem uses a simple reaction network modelling $^{12}\mathrm{C} + {}^{12}\mathrm{C} \rightarrow {}^{24}\mathrm{Mg}$ that can be integrated on the GPUs using the CUDA Fortran port of VODE \citep{vode} described in \cite{Zingale2018}.
Both tests were run on a single node, consisting of two IBM POWER9 processors (together providing 42 physical CPU cores) and six NVIDIA Volta GPUs, and compiled using the PGI compiler version 19.4. The CPU test used 42 MPI processes, each with 4 OpenMP threads, and the GPU test used 6 MPI processes, each process associated with a single GPU.
Note that the plot's $x$-axis scale is logarithmic. As can be seen, large speedups in excess of 30-40$\times$ were achieved for several of the functions when run on the GPU. Once we have finished porting the remaining routines to GPU and further optimised our code, we hope to achieve this sort of speedup for the entire timestep.
